---
title: "Conditional probability"
author: ""
date: ""
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r global_options, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      comment = "#>", highlight = TRUE,
                      fig.align = "center")
```

## Main ideas

- Use formulas to compute conditional probabilities from tabular data

- Compute empirical probabilities in R via simulation

## Packages

```{r packages}
library(tidyverse)
library(vcd) # used for Arthritis data
```

## Computing conditional probabilities

```{r data}
data(Arthritis)
glimpse(Arthritis)
```

We'll again work with `Arthritis` as we did last lecture. 

Let's look at the data in a tabular view. Don't worry about understanding these
functions, we're only using it to better visualize our data via a table.

```{r tabular_data}
xtabs(~ Treatment + Improved, data = Arthritis) %>% 
  addmargins()
```

- What is the probability a randomly selected patient received the placebo
  and had a marked improvement?
  
  $P(L \cap M) = 7 / 84$
  
- What is the probability a randomly selected patient had a marked improvement
  given they received the placebo?
  
  $P(M | L) = P(M \cap L) / P(L) = 7 / 43$
  
- What is the probability a randomly selected patient had no improvement given
  they received the treatment?
  
  $P(N | T) = P(N \cap T) / P(T) = 13 / 41$
  
- What is the probability a randomly selected patient was on the placebo given
  they had a marked improvement?
  
  $P(L | M) = P(L \cap M) / P(M) = 7/28$
  
- What is the probability a randomly selected patient was on the treatment
  given they had some or marked improvement?
  
  $P(T | S \cup M) = P(T \cap (S \cup M)) / P(S \cup M) = 28 / 42$

## Monty Hall Problem

### Game set-up and assumptions

Watch the first 38 seconds of the first video on the site here:
https://brilliant.org/wiki/monty-hall-problem/.

"Suppose you're on a game show, and you're given the choice of three doors: 
Behind one door is a car; behind the others, goats. You pick a door, say No. 1,
and the host, who knows what's behind the doors, opens another door, say No. 3, 
which has a goat. He then says to you, "Do you want to pick door No. 2?" Is it 
to your advantage to switch your choice?"

We will investigate the above decision of whether to switch or not to switch.

Assumptions:

1. The host will always open a door not picked by the contestant.

2. The host will always open a door which reveals a goat (i.e. not a car).

3. The host will always offer the contestant the chance to switch to 
   another door.
   
4. The door behind which the car is placed is chosen at random.

5. The door initially chosen by the contestant is chosen at random.


```{r doors}
doors <- c(1, 2, 3)
```

```{r set_up_selection}
monty_hall <- tibble(
  car_door   = sample(doors, size = 10000, replace = TRUE),
  my_door    = sample(doors, size = 10000, replace = TRUE)
  )

monty_hall
```


```{r monty_door}
monty_hall <- monty_hall %>% 
  rowwise() %>% 
  mutate(monty_door = if_else(car_door == my_door,
                              sample(doors[-my_door], size = 1),
                              6 - (car_door + my_door))) %>% 
  ungroup()

monty_hall
```

```{r stay_switch_win}
monty_hall <- monty_hall %>% 
  mutate(switch_win = car_door != my_door,
         stay_win   = car_door == my_door)

monty_hall
```

```{r estimate_probabilities}
monty_hall %>% 
  summarise(switch_win_prob = mean(switch_win),
            stay_win_prob   = mean(stay_win))
```

### Backstory

>You blew it, and you blew it big! Since you seem to have difficulty grasping
the basic principle at work here, I'll explain. After the host reveals a goat, 
you now have a one-in-two chance of being correct. Whether you change your 
selection or not, the odds are the same. There is enough mathematical 
illiteracy in this country, and we don't need the world's highest IQ 
propagating more. Shame!
<br><br>
**Scott Smith, Ph.D.**
**University of Florida**

>May I suggest that you obtain and refer to a standard textbook on probability 
before you try to answer a question of this type again?
<br><br>
**Charles Reid, Ph.D.**
**University of Florida**

These are just two of over 10,000 letters Marilyn vos Savant received in regards
to her response of the Monty Hall Problem. What was her response? "Yes; you 
should switch," she replied. "The first door has a 1/3 chance of winning, 
but the second door has a 2/3 chance."

Read more about this at 
https://priceonomics.com/the-time-everyone-corrected-the-worlds-smartest/

# References

1. https://priceonomics.com/the-time-everyone-corrected-the-worlds-smartest/
2. https://brilliant.org/wiki/monty-hall-problem/